### 2020-09-12 code/supplemental_v2.Rmd

```{r supplementary-figure-s1, fig.width = 9}
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data

@@

par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(2,2,0,0))
for(n in cancer_type[1:6]) {
with(umap_cancertype %>% 
       filter(project_code == n),
     scatter3D(plot_y, -plot_z, plot_x, 
                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
title(xlab="UMAP 2", line=0)
legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
}

par(mfrow = c(2,3), mai = c(0.1, 0.1, 2, 0.1))
par(mar = c(2,2,0,0))
for(n in cancer_type[7:12]) {
with(umap_cancertype %>%
       filter(project_code == n),
     scatter3D(plot_y, -plot_z, plot_x, 
                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
title(xlab="UMAP 2", line=0)
legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
}
par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(2,2,0,0))
for(n in cancer_type[13:15]) {
with(umap_cancertype %>%
       filter(project_code == n),
     scatter3D(plot_y, -plot_z, plot_x, 
                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
title(xlab="UMAP 2", line=0)
legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
}

par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(2,2,0,0))
for(n in cancer_type[19:23]) {
with(umap_cancertype %>%
       filter(project_code == n),
     scatter3D(plot_y, -plot_z, plot_x,
                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
title(xlab="UMAP 2", line=0)
legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
}

```
**Supplementary figure S1. Distribution of cancer type on UMAP clusters.** Each panel shows the identity of 
individual tumors by one of 23 cancer types projected onto the cluster UMAP identified 
in **Figure 1B of the main text**.

@@

```{r supplementary-figure-s2-data-generation, fig.width = 9}
ditto_color <- c(
  "#E69F00", "#56B4E9", "#009E73", "#F0E442",
  "#0072B2", "#D55E00", "#CC79A7", "#666666",
  "#AD7700", "#1C91D4", "#007756", "#D5C711",
  "#005685", "#A04700", "#B14380", "#4D4D4D",
  "#FFBE2D", "#80C7EF", "#00F6B3", "#F4EB71",
  "#06A5FF", "#FF8320", "#D99BBD", "#8C8C8C",
  "#FFCB57", "#9AD2F2", "#2CFFC6", "#F6EF8E",
  "#38B7FF", "#FF9B4D", "#E0AFCA", "#A3A3A3",
  "#8A5F00", "#1674A9", "#005F45", "#AA9F0D",
  "#00446B", "#803800", "#8D3666", "#3D3D3D")

cancer_w_subtypes <- list()
for(n in cancer_type) {
  spec_cancer <- umap_cancertype %>% 
    mutate(J_color = case_when(project_code == n ~ Subtype_Selected,
                               TRUE ~ "Other")) %>%
    filter(J_color != "Other")
  if(nrow(spec_cancer) == 0) {next
  }
  cancer_w_subtypes[[n]] <- n
}

@@

par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(2,2,0,0))

for(n in cancer_w_subtypes[1:6]) {
  spec_cancer <- umap_cancertype %>% 
    mutate(J_color = case_when(project_code == n ~ Subtype_Selected,
                               TRUE ~ "Other")) %>%
    filter(J_color != "Other")
  if(nrow(spec_cancer) == 0) {next
    }
  subtypes <- unique(spec_cancer$J_color)
  J_colors <- c(ditto_color[c(2:(length(subtypes) + 1))])#, 'grey')
  names(J_colors) <- subtypes

  with(spec_cancer,
     scatter3D(plot_y, -plot_z, plot_x, 
               bg = J_colors[J_color], pch = 21, cex = 0.8, lwd = 0.2,
               theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
               colvar = NULL))
  legend("top", inset = c(0.75), legend=n, bty = "n", horiz = T, cex = 1.2)
  legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
  legend("top", inset = c(0.10), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
  legend("top", inset = c(0.20), legend=names(J_colors[5:6]), col=J_colors[5:6], pch = 20, bty = "n", horiz = T, cex = 1.2)
  title(xlab="UMAP 2", line=0)
}

par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(2,2,0,0))

for(n in cancer_w_subtypes[7:12]) {
  spec_cancer <- umap_cancertype %>% 
    mutate(J_color = case_when(project_code == n ~ Subtype_Selected,
                               TRUE ~ "Other")) %>%
    filter(J_color != "Other")
  if(nrow(spec_cancer) == 0) {next
    }
  subtypes <- unique(spec_cancer$J_color)
  J_colors <- c(ditto_color[c(2:(length(subtypes) + 1))])#, 'grey')
  names(J_colors) <- subtypes

  with(spec_cancer,
     scatter3D(plot_y, -plot_z, plot_x, 
               bg = J_colors[J_color], pch = 21, cex = 0.8, lwd = 0.2,
               theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
               colvar = NULL))
  legend("top", inset = c(0.75), legend=n, bty = "n", horiz = T, cex = 1.2)
  legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
  legend("top", inset = c(0.10), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
  legend("top", inset = c(0.20), legend=names(J_colors[5:6]), col=J_colors[5:6], pch = 20, bty = "n", horiz = T, cex = 1.2)
  title(xlab="UMAP 2", line=0)
}

par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(2,2,0,0))

for(n in cancer_w_subtypes[13:15]) {
  spec_cancer <- umap_cancertype %>% 
    mutate(J_color = case_when(project_code == n ~ Subtype_Selected,
                               TRUE ~ "Other")) %>%
    filter(J_color != "Other")
  if(nrow(spec_cancer) == 0) {next
    }
  subtypes <- unique(spec_cancer$J_color)
  J_colors <- c(ditto_color[c(2:(length(subtypes) + 1))])#, 'grey')
  names(J_colors) <- subtypes

  with(spec_cancer,
     scatter3D(plot_y, -plot_z, plot_x, 
               bg = J_colors[J_color], pch = 21, cex = 0.8, lwd = 0.2,
               theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
               colvar = NULL))
  legend("top", inset = c(0.75), legend=n, bty = "n", horiz = T, cex = 1.2)
  legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
  legend("top", inset = c(0.10), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
  legend("top", inset = c(0.20), legend=names(J_colors[5:6]), col=J_colors[5:6], pch = 20, bty = "n", horiz = T, cex = 1.2)
  title(xlab="UMAP 2", line=0)
}

par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(2,2,0,0))

for(n in cancer_w_subtypes[16:19]) {
  spec_cancer <- umap_cancertype %>% 
    mutate(J_color = case_when(project_code == n ~ Subtype_Selected,
                               TRUE ~ "Other")) %>%
    filter(J_color != "Other")
  if(nrow(spec_cancer) == 0) {next
    }
  subtypes <- unique(spec_cancer$J_color)
  J_colors <- c(ditto_color[c(2:(length(subtypes) + 1))])#, 'grey')
  names(J_colors) <- subtypes

  with(spec_cancer,
     scatter3D(plot_y, -plot_z, plot_x, 
               bg = J_colors[J_color], pch = 21, cex = 0.8, lwd = 0.2,
               theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
               colvar = NULL))
  legend("top", inset = c(0.75), legend=n, bty = "n", horiz = T, cex = 1.2)
  legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1.2)
  legend("top", inset = c(0.10), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1.2)
  legend("top", inset = c(0.20), legend=names(J_colors[5:6]), col=J_colors[5:6], pch = 20, bty = "n", horiz = T, cex = 1.2)
  title(xlab="UMAP 2", line=0)
}

# ggplot(umap_cancertype, aes(x = plot_y, y = plot_x, colour = Subtype_Selected)) + 
#   geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
#   geom_point(size = 0.25) +
#   theme_classic() +
#   coord_fixed() +
#   xlab("UMAP 1") + ylab("UMAP 2") +
#   theme(axis.text.x = element_blank(),
#     axis.text.y = element_blank(),
#     axis.title.x = element_blank(),
#     axis.title.y = element_blank(),
#     axis.ticks = element_blank(),
#     axis.line = element_blank(),
#     legend.position = "none") +
#   facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
```
**Supplementary Figure S2. Histological subtypes.** Each panel shows the histological subtypes of one of the 23 
cancers projected onto the tumors of the UMAP from **Figure 1B of the main text**.

@@

# ##oncogene
# oncogene <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","MDM2","MYC","PIK3CA","TET2","USP6") ## list of selected oncogene
# ##tumor_supressor
# tumor_suppressor <- c("APC","BRIP1","CDKN2C","CHEK2","MSH2","PTEN","RB1","SMARCA4","TP53","VHL", "APOBEC3B") ## list of selected tumor suppressor
# 
# selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% c(oncogene, tumor_suppressor)) ## getting ensembl_id for the oncogenes for filtering gdc data

@@

selected_mut <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y, plot_z) %>% ## select relevant columns
  left_join(., selected_mut, by = "sample_id") %>%
  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y", "plot_z")) %>%
  filter(value != 0)

@@

```{r supplementary-figure-s3, fig.width = 9}
# out.width = '77%'
load("../data/selected_mut.Rda")

@@

par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(2,2,0,0))
for(n in genes[1:6]) {
with(selected_mut %>% 
       filter(variable == n),
     scatter3D(plot_y, -plot_z, plot_x, 
                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
title(xlab="UMAP 2", line=0)
legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
}

par(mfrow = c(2,3), mai = c(0.1, 0.1, 2, 0.1))
par(mar = c(2,2,0,0))
for(n in genes[7:12]) {
with(selected_mut %>% 
       filter(variable == n),
     scatter3D(plot_y, -plot_z, plot_x, 
                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
title(xlab="UMAP 2", line=0)
legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
}
par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(2,2,0,0))
for(n in genes[13:15]) {
with(selected_mut %>% 
       filter(variable == n),
     scatter3D(plot_y, -plot_z, plot_x, 
                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
title(xlab="UMAP 2", line=0)
legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
}
```
**Supplementary Figure S3. Distribution of mismatch repair genes on UMAP clusters.** Each panel shows one of 15 
genes involved in genome stability projected on the UMAP from **Figure 1B of the main text**.

@@

```{r supplementary-figure-s4-data-generation, eval = FALSE}
load("../data/gdc_reactome_path_analysis.Rda") ## load data
umap_cancertype <- fread("../data/umap_3d_coors.tsv") 
n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway

toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway

selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "sample_id") ## prepare pathway for left_join

selected_path <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y, plot_z) %>% ## select relevant columns
  left_join(., selected_path, by = "sample_id") %>%
  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y", "plot_z")) %>%
  filter(value != 0)

save(selected_path, file = "../data/selected_path.Rda")

# load("../data/selected_path.Rda") ## load data
# load("../data/gdc_reactome_path_analysis.Rda")
# umap_cancertype <- fread("../data/umap_3d_coors.tsv")
# umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet
# 
# n_path <- 377 ## number of pathways we want to see
# 
# toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway.. needed for facet
# 
# ## creating layout to plot all 377 pathways
# figs4_layout <- data.frame(start = seq(from = 1, to = length(toppath), by = 25),
#                           end = c(seq(from = 25, to = length(toppath), by = 25), length(toppath)))
# 
# for(n in 1:nrow(figs4_layout)) {
#   message(n)
#   pathway_layout <- figs4_layout %>% .[n,] ## getting the layout for selecting pathways
#   pathway_layout <- toppath[pathway_layout$start:pathway_layout$end] ## filtering those pathways from toppath
#   filtered_path <- selected_path %>% filter(variable %in% pathway_layout) %>% ## filtering our data to only include those pathways
#     mutate(variable = factor(variable, levels = c(pathway_layout)))
#   ggplot_umap <- ggplot(filtered_path, aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
#     geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
#     geom_point(size = 0.25) +
#     theme_classic() +
#     coord_fixed() +
#     xlab("UMAP 1") + ylab("UMAP 2") +
#     scale_colour_manual(values = rep("firebrick3", 25)) +
#     theme(axis.text.x = element_blank(),
#       axis.text.y = element_blank(),
#       axis.title.x = element_blank(),
#       axis.title.y = element_blank(),
#       axis.ticks = element_blank(),
#       axis.line = element_blank(),
#       strip.text = element_text(size = 5),
#       legend.position = "none") +
#     facet_wrap(~ variable,
#                ncol = 5) ## facet by pathways
#   new_dfname <- paste0("ggplot_umap_", n)
#   assign(paste0(new_dfname), value = ggplot_umap)
# }
# 
# save(ggplot_umap_1,ggplot_umap_2,ggplot_umap_3,ggplot_umap_4,ggplot_umap_5,ggplot_umap_6,ggplot_umap_7,ggplot_umap_8,ggplot_umap_9,ggplot_umap_10,ggplot_umap_11,ggplot_umap_12,ggplot_umap_13,ggplot_umap_14,ggplot_umap_15,ggplot_umap_16, file = "../data/fig_s4_ggplot.Rda")
```

```{r figure-s4-plot, fig.width= 9}
load("../data/selected_path.Rda")
load("../data/effect_heatmap.Rda") ## load effect heatmap data
pathorder <- as.character(levels(effect_heatmap$path_name)) ## getting col order

startingpoint <- seq(1, 377, by = 15)
startingpointlist <- list()
for(n in startingpoint) {
  start1 <- n
  end1 <- start1 + 5
  start2 <- end1 + 1
  end2 <- start2 + 5
  start3 <- end2 + 1
  end3 <- start3 + 2
  startingpointlist[[n]] <- data.frame(start1 = start1,
                                       end1 = end1,
                                       start2 = start2,
                                       end2 = end2,
                                       start3 = start3,
                                       end3 = end3)
}
startingpointlist <- as.data.frame(do.call(rbind, startingpointlist))

s4plot <- function(x) {
  a <-  x[,"start1"]
  b <- x[,"end1"]
  c <- x[,"start2"]
  d <- x[,"end2"]
  e <- x[,"start3"]
  f <- x[,"end3"]
  path01<- pathorder[a:b]
  path02<- pathorder[c:d]
  path03<- pathorder[e:f]
par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(2,2,0,0))
for(n in path01) { #pathorder[x[,"start1"]:x[,"end1"]]
  pathname <- as.data.frame(strsplit(n, " ")) ## split string into word
  pathname <- pathname[,1] ## get the character vectors
  with(selected_path %>% 
       filter(variable == n),
     scatter3D(plot_y, -plot_z, plot_x, 
                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
title(xlab="UMAP 2", line=0)
path1 <- paste0(pathname[1], " ", pathname[2], " ", pathname[3], " ", pathname[4], " ")
path1 <- gsub("NA", "", path1)
path2 <- paste0(pathname[5], " ", pathname[6], " ", pathname[7], " ", pathname[8], " ")
path2 <- gsub("NA", "", path2)
legend("top", inset = c(0.01), legend=path1, bty = "n", horiz = T, cex = 0.9)
legend("top", inset = c(0.06), legend=path2, bty = "n", horiz = T, cex = 0.9)
}

par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(2,2,0,0))
for(n in pathorder[x[,"start2"]:x[,"end2"]]) {
  pathname <- as.data.frame(strsplit(n, " ")) ## split string into word
  pathname <- pathname[,1] ## get the character vectors
  with(selected_path %>% 
       filter(variable == n),
     scatter3D(plot_y, -plot_z, plot_x, 
                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
title(xlab="UMAP 2", line=0)
path1 <- paste0(pathname[1], " ", pathname[2], " ", pathname[3], " ", pathname[4], " ")
path1 <- gsub("NA", "", path1)
path2 <- paste0(pathname[5], " ", pathname[6], " ", pathname[7], " ", pathname[8], " ")
path2 <- gsub("NA", "", path2)
legend("top", inset = c(0.01), legend=path1, bty = "n", horiz = T, cex = 0.9)
legend("top", inset = c(0.06), legend=path2, bty = "n", horiz = T, cex = 0.9)
}

par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(2,2,0,0))
for(n in pathorder[x[,"start3"]:x[,"end3"]]) {
  pathname <- as.data.frame(strsplit(n, " ")) ## split string into word
  pathname <- pathname[,1] ## get the character vectors
  with(selected_path %>% 
       filter(variable == n),
     scatter3D(plot_y, -plot_z, plot_x, 
                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
title(xlab="UMAP 2", line=0)
path1 <- paste0(pathname[1], " ", pathname[2], " ", pathname[3], " ", pathname[4], " ")
path1 <- gsub("NA", "", path1)
path2 <- paste0(pathname[5], " ", pathname[6], " ", pathname[7], " ", pathname[8], " ")
path2 <- gsub("NA", "", path2)
legend("top", inset = c(0.01), legend=path1, bty = "n", horiz = T, cex = 0.9)
legend("top", inset = c(0.06), legend=path2, bty = "n", horiz = T, cex = 0.9)
}
}

data <- startingpointlist[1,]
s4plot(x = data)
data <- startingpointlist[2,]
s4plot(x = data)
data <- startingpointlist[3,]
s4plot(x = data)
data <- startingpointlist[4,]
s4plot(x = data)
data <- startingpointlist[5,]
s4plot(x = data)
data <- startingpointlist[6,]
s4plot(x = data)
data <- startingpointlist[7,]
s4plot(x = data)
data <- startingpointlist[8,]
s4plot(x = data)
data <- startingpointlist[9,]
s4plot(x = data)
data <- startingpointlist[10,]
s4plot(x = data)
data <- startingpointlist[11,]
s4plot(x = data)
data <- startingpointlist[12,]
s4plot(x = data)
data <- startingpointlist[13,]
s4plot(x = data)
data <- startingpointlist[14,]
s4plot(x = data)
data <- startingpointlist[15,]
s4plot(x = data)
data <- startingpointlist[16,]
s4plot(x = data)
data <- startingpointlist[17,]
s4plot(x = data)
data <- startingpointlist[18,]
s4plot(x = data)
data <- startingpointlist[19,]
s4plot(x = data)
data <- startingpointlist[20,]
s4plot(x = data)
data <- startingpointlist[21,]
s4plot(x = data)
data <- startingpointlist[22,]
s4plot(x = data)
data <- startingpointlist[23,]
s4plot(x = data)
data <- startingpointlist[24,]
s4plot(x = data)
data <- startingpointlist[25,]
s4plot(x = data)
```

```{r, fig.width = 6.7}
par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(2,2,0,0), cex.lab = 0.8)
for(n in pathorder[376:377]) { 
  pathname <- as.data.frame(strsplit(n, " ")) ## split string into word
  pathname <- pathname[,1] ## get the character vectors
  with(selected_path %>% 
       filter(variable == n),
     scatter3D(plot_y, -plot_z, plot_x, 
                                bg = "grey", pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
title(xlab="UMAP 2", line=-1.7)
path1 <- paste0(pathname[1], " ", pathname[2], " ", pathname[3], " ", pathname[4], " ")
path1 <- gsub("NA", "", path1)
path2 <- paste0(pathname[5], " ", pathname[6], " ", pathname[7], " ", pathname[8], " ")
path2 <- gsub("NA", "", path2)
legend("top", inset = c(0.11), legend=path1, bty = "n", horiz = T, cex = 0.8)
legend("top", inset = c(0.16), legend=path2, bty = "n", horiz = T, cex = 0.8)
}
```

**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.** Each panel shows one of 377 Reactome
pathways projected on the UMAP from **Figure 1D of the main text** ordered by the hierarchical pathway order 
(column) from **Figure 1B of the main text**.

```{r, supplemental-s5-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data

@@

s5_data <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y, plot_z) %>% ## select relevant columns
  left_join(., selected_mut, by = "sample_id") %>%
  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y", "plot_z")) %>%
  filter(value != 0)

@@

```{r supplemental-s5-plot, fig.width=9}
load("../data/s5_data.Rda")
genes <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","PIK3CA","CDKN2C","PTEN","RB1","SMARCA4","TP53","VHL")

>>>>

\pagenumbering{gobble}

```{r supplementary-figure-s1, fig.width = 9, fig.height = 10}
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
load("../data/kelly.colours.rda")
cluster.colours <- c(kelly.colours[c(3:12)]) ## Color scheme
names(cluster.colours) <- c(1:10) ## Assign cluster to color

@@

par(mfrow = c(6,4), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(0,0,0,0))
for(n in cancer_type[1:23]) {
with(umap_cancertype %>% 
       filter(project_code == n),
     scatter3D(plot_y, -plot_z, plot_x, 
                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F, axes = FALSE,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
legend("bottom", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2, adj = 0.25)
}
```

**Supplementary figure S1. Distribution of cancer type on UMAP clusters.** Each panel shows the identity of 
individual tumors by one of 23 cancer types projected onto the cluster UMAP identified 
in **Figure 1B of the main text**. The colors indicate the tumor's cluster identity.

@@

```{r supplementary-figure-s2-data-generation, fig.width = 9, fig.height = 10}
load("../data/ditto.colours.rda")

## Determining which cancer type has subtype information
cancer_w_subtypes <- list()
for(n in cancer_type) {
  spec_cancer <- umap_cancertype %>% 
    mutate(J_color = case_when(project_code == n ~ Subtype_Selected,
                               TRUE ~ "Other")) %>%
    filter(J_color != "Other")
  if(nrow(spec_cancer) == 0) {next
  }
  cancer_w_subtypes[[n]] <- n
}

@@

par(mfrow = c(6,4), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(0,0,0,0))

for(n in cancer_w_subtypes[1:19]) {
  spec_cancer <- umap_cancertype %>% 
    mutate(J_color = case_when(project_code == n ~ Subtype_Selected,
                               TRUE ~ "Other")) %>%
    filter(J_color != "Other")
  if(nrow(spec_cancer) == 0) {next
    }
  subtypes <- unique(spec_cancer$J_color)
  J_colors <- c(ditto_colours[c(2:(length(subtypes) + 1))])#, 'grey')
  names(J_colors) <- subtypes

  with(spec_cancer,
     scatter3D(plot_y, -plot_z, plot_x, 
               bg = J_colors[J_color], pch = 21, cex = 0.8, lwd = 0.2,
               theta = 0, phi = 65, scale = F, axes = FALSE,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
               xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
               colvar = NULL))
  legend("bottom", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2, adj = 0.25)
  legend("top", inset = c(0), legend=names(J_colors[1:2]), col=J_colors[1:2], pch = 20, bty = "n", horiz = T, cex = 1)
  legend("top", inset = c(0.08), legend=names(J_colors[3:4]), col=J_colors[3:4], pch = 20, bty = "n", horiz = T, cex = 1)
  legend("top", inset = c(0.16), legend=names(J_colors[5:6]), col=J_colors[5:6], pch = 20, bty = "n", horiz = T, cex = 1)
}
```

**Supplementary Figure S2. Histological subtypes.** Each panel shows the histological subtypes of one of the 23 
cancers projected onto the tumors of the UMAP from **Figure 1B of the main text**. The colors indicate the tumor's histological subtype identity.

@@

@@

selected_mut <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y, plot_z, clust_knn) %>% ## select relevant columns
  left_join(., selected_mut, by = "sample_id") %>%
  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y", "plot_z", "clust_knn")) %>%
  filter(value != 0)

@@

```{r supplementary-figure-s3, fig.width = 9, fig.height = 10}
load("../data/selected_mut.Rda")

@@

par(mfrow = c(6,4), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(0,0,0,0))
for(n in genes[1:15]) {
with(selected_mut %>% 
       filter(variable == n),
     scatter3D(plot_y, -plot_z, plot_x, 
                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F, axes = FALSE,
                                xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
title(xlab="UMAP 2", line=0)
legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
}
```

**Supplementary Figure S3. Distribution of mismatch repair genes on UMAP clusters.** Each panel shows one of 15 
genes involved in genome stability projected on the UMAP from **Figure 1B of the main text**. The colors indicate the tumor's cluster identity.

@@

```{r, supplemental-s4-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data

@@

s5_data <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y, plot_z, clust_knn) %>% ## select relevant columns
  left_join(., selected_mut, by = "sample_id") %>%
  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y", "plot_z", "clust_knn")) %>%
  filter(value != 0)

@@

```{r supplemental-s4-plot, fig.width=9}
load("../data/s5_data.Rda")
genes <- c("AKT2","BCL2","CTNNB1","ERBB2","KRAS","PIK3CA","CDKN2C","PTEN","RB1","SMARCA4","TP53","VHL")

par(mfrow = c(2,3), mai = c(0.1, 0.1, 0.1, 0.1))
par(mar = c(2,2,0,0))
for(n in genes[1:6]) {
with(s5_data %>% 
       filter(variable == n),
     scatter3D(plot_y, -plot_z, plot_x, 
                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2,
                                theta = 0, phi = 65, scale = F,
               xlim = c(-6, 4.5), zlim = c(-4.5, 4), ylim = c(-4, 5),
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
title(xlab="UMAP 2", line=0)
legend("top", inset = c(0), legend=n, bty = "n", horiz = T, cex = 1.2)
}
